Military mechanization as a complex adaptive system

### 2018-07-18 paper2/paper2.tex
good job tom

This paper is divided into five sections. This \textbf{Introduction} gives
background on military mechanization. The \textbf{Theory} section rigorously
develops mechanization as a complex adaptive system best represented by
agent-based modeling. The \textbf{Model} section presents a hypothesis and
implements an agent-based model to test that hypothesis.
\textbf{Results} are shown. Implications are explored in \textbf{Discussion}.

@@

with $y_{ij}$ the mechanization level of state $i$ in year $j$ (measured as the 
ratio of ground combat vehicles to infantry in a state's army) and $x_1 \ldots
x_j \ldots x_{P}$ covariates hypothesized to be associated with
mechanization. They conclude that, out of the
potential causes mentioned above, ``choices about
mechanization are strongly associated with a state's security environment.''
States that have engaged in conflict with a highly-mechanized adversary and states
with highly mechanized neighbors tend to higher mechanization; conversely, states facing
domestic insurgency or that have lost counter-insurgency wars tend to have lower
mechanization rates. Their dataset and conclusions have guided further scholarship on military
mechanization, e.g. \citet{pilster2012democracies}, which mostly follow the same
statistical approach.

@@

There is no single universally accepted definition of a complex adaptive system
\citep{miller2009complex}. However, some core features are
that the system is non-decomposable; that the system features autonomous, cognitive
agents; and that the individual behaviors of these agents aggregate upwards into
complex system behavior (e.g., stability, chaos, feedback loops, and the like). 
The following sections detail how mechanization meets the first two parts of
this description (later sections deal with system behavior).

@@

Figure 2 is the logic diagram resulting from the above. It must be emphasized that we are only scratching the 
surface of system complexity. For example, Figure 2 does not address the nature of relationships 
between covariates: linear or non-linear? subject to multi-way
interactions? necessity and sufficiency distinctions? Moreover, it only includes 
abstract variables and not individual agents (a point addressed in the following
sub-section). But even as-is, Figure 2 visually conveys the essential point: \textit{the system is
not tractable to decomposition}. We cannot isolate one component from the
greater whole.

This system complexity disallows previously-mentioned statistical techniques
for correction of selection bias. Both Heckman correction and propensity score
matching boil down to using covariates to predict (and hence control for) the chances 
of receiving treatment. For example, we might assess the effect of war on
mechanization by controlling for the factors that affect propensity for war. 
But these techniques break down if the confounding variables also directly
relate to the outcome of interest, or if the direction of causality flows both
wars. Both are true for mechanization 
(c.f. Figure 2). This ``post-treatment bias'' \citep{king2010hard} means we need
some technique other than econometrics to perform causal inference regarding
mechanization.

We would like to perform ``causal surgery'' \citep{pearl2009causality}, i.e. modify the value at one
of the nodes and observe the resultant system state. This sort of
experimentation might allow us observe the system's sensitivity (or lack thereof) to changes
in input. But, as previously covered, one generally cannot conduct
experimental trials in international relations.  A modeling paradigm that would let us do so is needed.

@@

Orthodox studies of mechanization incorporate cognition to some degree. For example,
\citet{sechser2010army} test for learning: whether states that fail in
counter-insurgency wars decrease their mechanization as a result of this
``anti-mechanization lesson.'' This depiction, however, fails to capture 
key phenomena.

Human beings learn not only from their own experiences, but also from observing others.
State decisionmakers are no exception. For example, the stunning success of
Coalition precision-guided munitions in the 1991 Gulf War led other states who
were not direct combatants to reform their own militaries
\citep{scobell2011chinese}. There is no guarantee that such emulation always
works out. Consider Ireland: between the 1920s and 1940s, it sought to spend a
large portion of its defense budget on building a small number of
tanks, which at that time were novel technology being implemented by the leading
European militaries. They would have been wholly insufficient to repel a British invasion (their ostensible purpose), 
and Ireland would have been better-served to pursue a low-mechanization guerrilla defense strategy, 
but tanks were seen as a prestigious signifier of a ``real'' professional
military and hence desirable
\citep{farrell1998professionalization,farrell2001transnational}.

@@

Complex systems and ABM have been used in IR. Many methodology-focused
works have identified their utility to the field \citep[some prominent examples
are][]{modelski1990world,jervis1998system,pepinsky2005agents,de2014agent}.
Although actual usage has somewhat lagged behind theoretical enthusiasm,
ABM has been used to explore emergent 
balance of power \citep{cederman1997emergent}, international negotiation
\citep{earnest2008coordination}, and international crises and conflict
\citep{masad2016dissertation}. It is especially popular for studying
population-centric security phenomena such as terrorism and insurgency
\citep{cioffi2010mason,bhavnani2012modeling,weidmann2013violence,bhavnani2014group}.

The author's sense is that adoption of ABM in IR has lagged because it is
regarded as non-empirical. ABM offers no guarantees: just
because one's model ``works'' (internally consistent, externally plausible) 
does not mean it is true. Econometrics are the dominant methodological approach 
in quantitative IR because they seem to offer an objective yardstick for competing hypotheses: 
does A or B best fit the data? This paper contends that
epistemological limitations (most importantly, selection bias stemming from
non-randomization, and system complexity) mean that econometrics do not
provide reliable statistical inference for certain classes of IR problems, among
them mechanization \citep[c.f.][for a concurring opinion]{king2010hard}. 
If neither econometrics nor ABM can offer 
certitude that a hypothesis is true---as is the case for mechanization---then the latter 
should be preferred for its advantages in experimentation, transparency, researcher control, 
and amenability to complex systems.

@@

This section of the paper describes research hypotheses and an agent-based
model used to test them.

\subsection{Research hypotheses}

I hypothesize that the arguments of \citet{sechser2010army} are incomplete and
that some omitted variable or set of variables acts as a ``brake'' on 
mechanization.\footnote{By ``omitted'', I mean that Sechser and Saunders conclude that the variable
is not very important for mechanization, not that they do not address the
variable at all.} An agent-based model is used to test this hypothesis. In broad
terms, agents (states) are defined to have the cognitive and learning behavior
implicitly claimed by \citet{sechser2010army}. The model is initialized with
real-world data from 1979 and allowed to run until 2001. The micro- and
macro-outcomes are then validated against real-world 2001 data. If
\citet{sechser2010army} are correct, the model's 2001 outcomes should match
real-world 2001 outcomes. If the model mechanization levels in 2001 are much
higher, then their claimed behavioral rules are incomplete and fail to include
some ``brake'' on mechanization, matching my hypothesis. Proving exactly what
that missing variable or set of variables is is outside the scope of this model
and paper, though it is discussed.

\subsection{Model}

@@

An important caveat is that the set of agents represented in the model does not
change over time: i.e., agents are the same from 1979 onwards. In real life,
of course, the agent population was not static over this time period, with many
states ceasing to exist in 1991 and new ones forming. Modeling these changes is
beyond the scope of this model. How a collapsed state's military assets are divvied up
among successor states is complex and not tractable to simple representation
(e.g. averaging). For example, the Soviet Union deployed most of its modern
military equipment in the far west of the country in anticipation of a conflict
with NATO, and so when the USSR collapsed Ukraine inherited the lion's share, leaving
little for successor states such as Azerbaijan that were not important military
districts \citep{roy1993military}. Such inheritance has
a strong effect on the mechanization of of successor states. Designing
complex and state-specific ``rules of inheritance'' was too much for this
initial effort but is an area for further development.

@@

@@

For some agents, however, mechanization levels are extremely high. 
Mechanization score is a percentage, and so values over 1 do not make sense, 
yet some countries end the simulation with mechanization scores over 1. Note
that one could easily right-bound values at 1, but this was not done in order to
see what values would be organically converged to.

@@

\section{Discussion}

@@

\subsection{Areas for Improvement}

Having taken a critical view of \citet{sechser2010army}, it only is fair that
this paper applies that lens to itself. There are two main areas for
improvement.

One, this model's representation of \citet{sechser2010army} could be tightened
up, as there currently is divergence between the two. For example, this model
examines a smaller agent set than their paper. Building ``inheritance''
processes to allow new states to emerge is desirable. This paper's
use of statistical coefficients from \citet{sechser2010army} also has some
methodological problems, a full description of which is omitted here due to
space constraints. It suffices to say that a better approach would be to conduct
more extensive parameter sweeps of relevant cognitive factors.

Two, this model could incorporate potential explanations of what the missing
``brake'' on mechanization might be. Economy and culture could be included in
the model. Doing so would allow exploration of whether these factors are
logically consistent and can act as a countervailing factor that controls the
positive feedback loop of mechanization. It cannot \textit{prove} that these or
any other factors actually exist in reality, but they can demonstrate how they
are or are not consistent in a simulated reality.

My hope is that making these changes will make this a publication-quality paper
capable of serving as one of two required papers for the comprehensive exams.

>>>>

This paper is divided into five sections. This \textbf{Introduction} gives
background on military mechanization. The \textbf{Theory} section rigorously
develops mechanization as a complex adaptive system best represented by
agent-based modeling. The \textbf{Model} section presents hypotheses and
implements an agent-based model to test them.
\textbf{Results} are shown. Implications are explored in \textbf{Discussion}.

@@

with $y_{ij}$ the mechanization level of state $i$ in year $j$ (measured as the 
ratio of ground combat vehicles to infantry in a state's army) and $x_1 \ldots
x_k \ldots x_{P}$ covariates hypothesized to be associated with
mechanization. They conclude that, out of the
potential causes mentioned above, ``choices about
mechanization are strongly associated with a state's security environment.''
States that have engaged in conflict with a highly-mechanized adversary and states
with highly mechanized neighbors tend to higher mechanization; conversely, states facing
domestic insurgency or that have lost counter-insurgency wars tend to have lower
mechanization rates. Their dataset and conclusions have guided further scholarship on military
mechanization \citep[e.g.][]{pilster2012democracies},

@@

There is no single universally accepted definition of a complex adaptive system
\citep{miller2009complex}. However, some core features are
that the system comprises many constituent parts that interact to form a
non-decomposable whole; that the system features autonomous, cognitive
agents; and that the relatively simple interactions of these agents aggregate upwards into
complex system behavior (e.g., stability, chaos, feedback loops, and the like). 
The following sections detail how mechanization meets the first two parts of
this description (later sections deal with system behavior).

@@

Figure 2 is the logic diagram resulting from the above. It must be emphasized that we are only scratching the 
surface of system complexity. For example, Figure 2 does not address the nature of relationships 
between covariates: linear or non-linear? Subject to multi-way
interactions? Necessity vs. sufficiency distinctions? Moreover, it only includes 
abstract variables and not individual agents (a point addressed in the following
sub-section). But even as-is, Figure 2 visually conveys the essential point: \textit{the system is
not tractable to decomposition}. We cannot isolate one component from the
greater whole. Separating cause from effect in such a system is challenging even
at the most basic conceptual level
\citep{pearl2009causality}. It also becomes statistically intractable.

This system complexity disallows previously-mentioned
techniques for mitigating selection bias. Both Heckman correction and propensity score
matching boil down to using covariates to predict (and hence control for) the chances 
of receiving treatment. For example, we might assess the effect of war on
mechanization by controlling for the factors that affect propensity for war. 
But these techniques break down if the confounding variables also directly
relate to the outcome of interest, or if the direction of causality flows both
ways. Both are true for mechanization 
(c.f. Figure 2). This ``post-treatment bias'' \citep{king2010hard} means we need
some technique other than econometrics to perform causal inference regarding
mechanization.

Complex systems theory \citep[ultimately stemming to][]{simon1996artificial}
offers a way forward. Although we cannot separate the many parts from the whole,
we would like to modify the value at one of the nodes and observe the resultant system state. This sort of
experimentation might allow us observe the system's sensitivity (or lack thereof) to changes
in input. But, as previously covered, one generally cannot conduct
experimental trials in international relations.  A modeling paradigm that would let us do so is needed.

@@

Moreover, human beings learn not only from their own experiences, but also from observing others.
State decisionmakers are no exception. For example, the stunning success of
Coalition precision-guided munitions in the 1991 Gulf War led other states who
were not direct combatants to reform their own militaries
\citep{scobell2011chinese}. There also is no guarantee that such cognition
follows typical assumptions of bounded rationality. 
Consider Ireland: between the 1920s and 1940s, it sought to spend a
large portion of its defense budget on building a small number of
tanks. This was a near-suicidal defense posture, given Ireland's
conventional weakness relative to the United Kingdom (the main object of Irish
defense thought). Rationally, Ireland should have pursued a low-mechanization,
guerrilla defense strategy given their weakness and given that exactly such a
strategy defeated the British in the 1919-22 war of independence. However,  
tanks at the time were novel technology being implemented by the leading
European militaries, and so were seen as a prestigious signifier of a ``real'' professional
military \citep{farrell1998professionalization,farrell2001transnational}. These examples
support the larger point that perception and cognition are fundamental to
understanding mechanization. A modeling framework that explicitly represents these factors
thus would be useful.

@@

Complex systems and ABM have been used in IR. Many methodology-focused
works have identified their utility to the field \citep[some prominent examples
are][]{modelski1990world,jervis1998system,pepinsky2005agents,de2014agent}.
Although actual usage has somewhat lagged behind theoretical
enthusiasm---a point addressed in more detail below---ABM has been used to explore emergent 
balance of power \citep{cederman1997emergent}, international negotiation
\citep{earnest2008coordination}, and international crises and conflict
\citep{masad2016dissertation}. It is especially popular for studying
population-centric security phenomena such as terrorism and insurgency
\citep{cioffi2010mason,bhavnani2012modeling,weidmann2013violence,bhavnani2014group}.

The author's sense is that adoption of ABM in IR has lagged because it is
regarded as non-empirical. ABM offers no guarantees: just
because one's model ``works''  does not mean it is true. Econometrics dominate quantitative 
IR because they offer an objective yardstick for competing hypotheses. 
This paper has demonstrated that econometrics, although useful in many situation, do not
provide reliable statistical inference for certain classes of IR problems \citep[c.f.][for a concurring opinion]{king2010hard}. 
For such cases (and mechanization is one of them), ABM no longer is at the
disadvantage in causal inference, and has important advantages in experimentation, 
researcher control, and amenability to complex systems

@@

This section of the paper describes the research strategy and specific
hypotheses, and an agent-based model used to test them.

\subsection{Research strategy and hypotheses}

This paper's strategy is not to propose a master
model of mechanization that incorporates all factors raised in Figure 2 and
the subsequent discussion. Rather, it is to start with a relatively simple
agent-based model of military mechanization that ``agentizes'' extant
traditional scholarship on mechanization. This research strategy is deliberate. 
The goal is to maintain maximum correspondence with extant scholarship on mechanization 
so that the results of an agentized version of it can be fairly compared.

The primary hypothesis is that the arguments of \citet{sechser2010army}, which
center on security environment, are incomplete; some omitted variable or set of
variables must act 
as a ``brake'' on  mechanization.\footnote{By ``omitted'', I mean that Sechser and Saunders conclude that the variable
is not very important for mechanization, not that they do not address the
variable at all.} If mechanization is externally-driven, then one agent raising their
mechanization causes another agent to do so, causes another agent to do so\ldots
and so on. This \textit{positive feedback loop} would reverbate through the
system and drive mechanization upward. Importantly, it acts constantly, whereas
the ``anti-mechanization lessons'' identified by \citet{sechser2010army} (e.g.,
defeat in counter-insurgency) are episodic. This paper thus hypothesizes that a
faithful agentization of this orthodox model will result in too-high
mechanization levels as compared to the historical record.

\subsection{Model overview}

An agent-based model is used to test this hypothesis. In broad
terms, agents (states) are defined to have the cognitive and learning behavior
implicitly claimed by \citet{sechser2010army}. The model is initialized with
real-world data from 1979 and allowed to run until 2001. The micro- and
macro-outcomes then are validated against real-world 2001 data. If
orthodox theory is correct, the model's 2001 outcomes should match
real-world 2001 outcomes. If the model mechanization levels in 2001 are much
higher, then their claimed behavioral rules are incomplete and fail to include
some ``brake'' on mechanization, matching the hypothesis.

@@

An important caveat is that the set of agents represented in the model does not
change over time: i.e., agents are the same from 1979 onwards. In real life,
of course, the agent population was not static over this time period, with many
states ceasing to exist in 1991 and new ones forming. Modeling these changes is
beyond the scope of this model. How a collapsed state's military assets are divvied up
among successor states is complex and not tractable to simple representation
(e.g. averaging). For example, the Soviet Union deployed most of its modern
military equipment in the far west of the country in anticipation of a conflict
with NATO, and so when the USSR collapsed Ukraine inherited the lion's share, leaving
little for successor states such as Azerbaijan that were not important military
districts \citep{roy1993military}. Such inheritance has
a strong effect on the mechanization of of successor states. Building a separate
sub-model for  complex ``rules of inheritance'' is beyond this initial effort.

@@

MORE ON VERIFICATION

@@

For some agents, however, mechanization levels are extremely high. 
Mechanization score is a ratio, and so values over 1 do not make sense, 
yet some countries end the simulation with mechanization scores over 1. Note
that one could easily right-bound values at 1, but this was not done in order to
see what values would be organically converged to.

@@

A parameter sweep reveals that this pattern is relatively insensitive
to input values. That is to say, changing \textit{NeighborsMechSensitivity}, 
\textit{EnemiesMechSensitivity}, and \textit{LessonSensitivity} alter the
magnitude of mechanization but not the overall trend of it being most pronounced
in states with many neighbors, enemies, and neighbor-enemies. For example, Table
6 shows the effect of lowering \textit{NeighborsMechSensitivity} and 
\textit{EnemiesMechSensitivity} to values lower than that used in the baseline
run. Doing so does not change what state ends up as the most mechanized at
simulation end, suggesting that the fundamental system behavior is unaltered.

\begin{table}[h]
	\centering
	\caption{Parameter sweep}
	\begin{tabular}{|l r|}
		\hline
		\textbf{Parameter Values} & \textbf{Most Mechanized State} \\ 
		1.0 (baseline) & Israel \\
		0.75 & Israel \\
		0.5 & Israel \\
		0.25 & Israel \\
		\hline
	\end{tabular}
\end{table}
\section{Discussion}

@@
